---
title: "ALA data analyst interview"
author: "Lu-Yi Wang"
date: "2023-02-28"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
subtitle: Techanical test
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---
View GitHub repo [here](https://github.com/luyiwangtw/Interview_ALA_data_analyst)

```{r Load R libraries, include = FALSE}
library(xaringanExtra) # for the clipboard
library(dplyr)
library(ggplot2)
library(ozmaps)
library(pander)
```

```{r Set up clipboard, include = FALSE}
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

Here, I select data including record type, spatial, temporal, and taxonomic information for the analyses.

```{r Import data}
dat.raw <- read.csv("../data/records-2023-02-28.csv", header = TRUE) %>% 
  select(basisOfRecord, year, month, day, decimalLatitude, decimalLongitude, scientificName, kingdom, phylum, class, order, family, genus, subgenus, species, taxonRank)
```

Additionally, I excluded material samples and preserved specimens to investigate the occurrence in the field.

```{r}
dat.raw %>% 
  count(basisOfRecord) %>% 
  rename("Record type" = "basisOfRecord", "Number of records" = "n") %>% 
  pander()
```

# Data overview 

```{r}
dat <- dat.raw %>% 
  filter(!grepl('MATERIAL_SAMPLE|PRESERVED_SPECIMEN', basisOfRecord)) %>% 
  mutate(season = case_when(               # create a season column
    month %in% c(12, 1, 2) ~ "summer",
    month %in% c(3, 4, 5) ~ "fall",
    month %in% c(6, 7, 8) ~ "winter",
    month %in% c(9, 10, 11) ~ "spring"))
```

```{r}
# There are blanks in the taxonomic columns that can not be removed as NA
dat.table <- dat %>%
  filter(!order == "" &
           !family == "" &
           !genus == "") 
```

In this dataset, 

1. the data span from year `r min(dat$year, na.rm = TRUE)` to `r max(dat$year, na.rm = TRUE)` (`r max(dat$year, na.rm = TRUE)-min(dat$year, na.rm = TRUE)+1` years in total).
2. there are `r length(unique(na.omit(dat.table$order)))` orders, `r length(unique(na.omit(dat.table$family)))` families, `r length(unique(na.omit(dat.table$genus)))` genera, and `r length(unique(na.omit(dat.table$species)))` species.
3. there are `r length(dat$basisOfRecord)` observations, but `r sum(is.na(dat$month))` of them do not have temporal information.

```{r}
tibble(Taxonomic_rank = c("order", "family", "genus", "species"),
       number = c(length(unique(na.omit(dat.table$order))), 
                  length(unique(na.omit(dat.table$family))), 
                  length(unique(na.omit(dat.table$genus))), 
                  length(unique(na.omit(dat.table$species))))) %>% 
  pander()
```

# Data exploration

## Occurance overtime {.tabset .tabset-fade .tabset-pills}

```{r}
dat$season <- factor(dat$season, levels = c("spring", "summer", "fall", "winter"))
dat$month <- factor(dat$month, levels = 1:12)
dat$year <- factor(dat$year)
```

### Monthly occurance

```{r}
dat %>%
  filter(!is.na(month)) %>% 
  group_by(month) %>%
  summarise(n = length(basisOfRecord)) %>% 
  ggplot(aes(x = month, y = n)) +
  geom_bar(stat = "identity", fill = "gold") +
  geom_text(aes(label = n), position = position_dodge(width = 0.9), vjust = -0.25) +
  labs(title = "Occurence over months",
       x = "Month", y = "Number of observations") +
  theme_classic() 
```

### Seasonal occurance

```{r}
dat %>%
  filter(!is.na(month)) %>% 
  group_by(season) %>%
  summarise(n = length(basisOfRecord)) %>% 
  ggplot(aes(x = season, y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = n), position = position_dodge(width = 0.9), vjust = -0.25) +
  labs(title = "Occurence over seasons",
       x = "Season", y = "Number of observations") +
  theme_classic() 
```

### Yearly occurance

```{r}
dat %>%
  filter(!is.na(year)) %>% 
  group_by(year) %>%
  summarise(n = length(basisOfRecord)) %>% 
  ggplot(aes(x = year, y = n)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  labs(title = "Occurence over years",
       x = "Year", y = "Number of observations") +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  theme_classic() 
```

## Maps {.tabset .tabset-fade .tabset-pills}

```{r}
oz_states <-  ozmaps::ozmap_states
dat.season <- dat %>% filter(!is.na(season)) # plot points that have season info
```



### Without Jervis Bay
```{r, warning=FALSE}
ggplot(oz_states) +
  geom_sf() +
  coord_sf(x = c(148.6, 149.5), y = c(-36, -35.1)) +
  geom_point(
    data = dat.season,
    aes(x = decimalLongitude, y = decimalLatitude, colour = season),
    alpha = 0.5) +
  xlab("Latitude") + 
  ylab("Longitude") +
  theme_bw()
```

### With Jervis Bay
```{r, warning=FALSE}
ggplot(oz_states) +
  geom_sf() +
  coord_sf(x = c(148.6, 150.8), y = c(-36, -35.1)) +
  geom_point(
    data = dat.season,
    aes(x = decimalLongitude, y = decimalLatitude, colour = season),
    alpha = 0.5) +
  xlab("Latitude") + 
  ylab("Longitude") +
  theme_bw()
```

### Only Jervis Bay

Some points on the sea. Might need to check the original data.
```{r, warning=FALSE}
ggplot(oz_states) +
  geom_sf() +
  coord_sf(x = c(150.55, 150.8), y = c(-35.22, -35.1)) +
  geom_point(
    data = dat.season,
    aes(x = decimalLongitude, y = decimalLatitude, colour = season),
    alpha = 0.5) +
  xlab("Latitude") + 
  ylab("Longitude") +
  theme_bw()
```

# Records over seasons

From the exploration above, I am interested in investigating the association between occurrence and season. 

Here, I would also like to look into if there is any difference in this pattern between families.

```{r, message=FALSE}
dat.season %>% 
  filter(!family == "") %>% 
  group_by(season, family) %>%
  summarise(n = length(basisOfRecord)) %>% 
  ggplot(aes(x = season, y = family, fill = log(n))) +
  geom_tile(color = "white", linewidth = .5, linetype = 1) +
  labs(title = "Reocrds in different families in different seasons",
       subtitle = "colour shows the record numbers",
       x = "Season", y = "Family") +
  theme_classic() 
```
_The blanks on the heatmap for Varanidae and Diplodactylidae due to the lack of records in winter._

_Number of records are natural log transformed to provide clearer visualisation._

From this figure, most families are observed more often during spring and summer.

# Spatial and temperal distribution

From the heatmap above, Scincidae, Elapidae and Agamidae have pretty seasonal seasonal records while Gekkonidae, Cheloniidae and Pythonidae have little seasonality. Here, I break it down into months to see their monthly occurrence.

## Seasonal records

```{r, warning=FALSE, message=FALSE}
dat.season %>% 
  filter(family %in% c("Scincidae", "Elapidae", "Agamidae", "Gekkonidae", "Cheloniidae", "Pythonidae") & !family == "") %>% 
  group_by(month, family) %>%
  summarise(n = length(basisOfRecord)) %>% 
  mutate(seasonality = case_when(
    family %in% c("Scincidae", "Elapidae", "Agamidae") ~ "seasonal",
    family %in% c("Gekkonidae", "Cheloniidae", "Pythonidae") ~ "constant")) %>% 
  ggplot(aes(x = month, y = family, fill = log(n))) +
  geom_tile(color = "white", linewidth = .5, linetype = 1) +
  facet_grid(seasonality ~., scales = "free", space = "free") +
  labs(title = "Reocrds over monthsin different families with varied seasonal pattern",
       subtitle = "colour shows the record numbers",
       x = "Season", y = "Family") +
  theme_classic() 
```

## Seasonal distribution {.tabset .tabset-fade .tabset-pills}

Other than temporal variation, I wonder if spatial distribution also varies with seasonality. Here, I use only the families selected from the above.

For convenience, here I exclude those in Jervis Bay.

```{r}
dat.season$month <- factor(dat.season$month, levels = c(12, 1:11))
```

### All families 

```{r, warning=FALSE}
dat.season$month <- factor(dat.season$month, levels = c(12, 1:11))
ggplot(oz_states) +
  geom_sf() +
  coord_sf(x = c(148.6, 149.5), y = c(-36, -35.1)) +
  scale_x_continuous(guide = guide_axis(angle = 90)) +
  geom_point(
    data = dat.season,
    aes(x = decimalLongitude, y = decimalLatitude, colour = season),
    alpha = 0.5) +
  facet_wrap(~ month, ncol = 6) +
  xlab("Latitude") + 
  ylab("Longitude") +
  theme_bw()
```

### Seasonal families {.tabset .tabset-fade .tabset-pills}

Season families expand their range during warmer seasons.       

#### Scincidae
```{r, warning=FALSE}
ggplot(oz_states) +
  geom_sf() +
  coord_sf(x = c(148.6, 149.5), y = c(-36, -35.1)) +
  scale_x_continuous(guide = guide_axis(angle = 90)) +
  geom_point(
    data = dat.season %>% filter(family == "Scincidae"),
    aes(x = decimalLongitude, y = decimalLatitude, colour = season),
    alpha = 0.5) +
  facet_wrap(~ month, ncol = 6) +
  xlab("Latitude") + 
  ylab("Longitude") +
  theme_bw()
```

#### Elapidae
```{r, warning=FALSE}
ggplot(oz_states) +
  geom_sf() +
  coord_sf(x = c(148.6, 149.5), y = c(-36, -35.1)) +
  scale_x_continuous(guide = guide_axis(angle = 90)) +
  geom_point(
    data = dat.season %>% filter(family == "Elapidae"),
    aes(x = decimalLongitude, y = decimalLatitude, colour = season),
    alpha = 0.5) +
  facet_wrap(~ month, ncol = 6) +
  xlab("Latitude") + 
  ylab("Longitude") +
  theme_bw()
```

#### Agamidae

```{r, warning=FALSE}
ggplot(oz_states) +
  geom_sf() +
  coord_sf(x = c(148.6, 149.5), y = c(-36, -35.1)) +
  scale_x_continuous(guide = guide_axis(angle = 90)) +
  geom_point(
    data = dat.season %>% filter(family == "Agamidae"),
    aes(x = decimalLongitude, y = decimalLatitude, colour = season),
    alpha = 0.5) +
  facet_wrap(~ month, ncol = 6) +
  xlab("Latitude") + 
  ylab("Longitude") +
  theme_bw()
```


### Constant families {.tabset .tabset-fade .tabset-pills}

Here I do not plot the family Cheloniidae and Pythonidae because they have relatively low record numbers (`r nrow(dat %>% filter(family == "Cheloniidae"))` and `r nrow(dat %>% filter(family == "Pythonidae"))` records, repectively).

From Gekkonidae, it looks like the distribution still quite restricted regardless of the season. However, this could also be because of the relatively low records (`r nrow(dat %>% filter(family == "Gekkonidae"))`) in this family and lower number of species (`r dat %>% filter(family == "Gekkonidae") %>% filter(!species == "") %>% distinct(species) %>% nrow()`) compared with seasonal families. For example, Scincidae has `r dat %>% filter(family == "Scincidae") %>% filter(!species == "") %>% distinct(species) %>% nrow()` species and `r nrow(dat %>% filter(family == "Scincidae"))` records.

#### Gekkonidae 
```{r, warning=FALSE}
ggplot(oz_states) +
  geom_sf() +
  coord_sf(x = c(148.6, 149.5), y = c(-36, -35.1)) +
  scale_x_continuous(guide = guide_axis(angle = 90)) +
  geom_point(
    data = dat.season %>% filter(family == "Gekkonidae"),
    aes(x = decimalLongitude, y = decimalLatitude, colour = season),
    alpha = 0.5) +
  facet_wrap(~ month, ncol = 6) +
  xlab("Latitude") + 
  ylab("Longitude") +
  theme_bw()
```